---
title: "Introducing Multiple Linear Regression"
author: "Dr. Maria Tackett"
date: "10.17.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta199-slides.css"
    logo: img/sta199-sticker-icon.png
    lib_dir: libs
    nature: 
      highlightLines: true
      highlightStyle: github
      countIncrementalSlides: false
      slideNumberFormat: "%current%"
---

layout: true

<div class="my-footer">
<span>
<a href="http://datasciencebox.org" target="_blank">datasciencebox.org</a>
</span>
</div> 

---

```{r packages, echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(knitr)
library(emo)
library(DT)
```

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE,
                      warning=FALSE,
                      message=FALSE,
                      fig.height = 2.65, 
                      dpi = 300) 
# For nonsese...
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
# For magick
dev.off <- function(){
  invisible(grDevices::dev.off())
}
```

```{r include=F}
library(tidyverse)
library(broom)
```

class: middle, center

### [Click for PDF of slides](08a-mlr-intro.pdf)

---

## Announcements

- Complete [Reading 05](https://www2.stat.duke.edu/courses/Fall19/sta199.001/reading/reading-05.html) (if you haven't already done so)

---

## Data & packages

```{r message=FALSE}
library(tidyverse)
library(broom)
```

```{r message=FALSE}
pp <- read_csv("data/paris_paintings.csv", 
               na = c("n/a", "", "NA"))
```

---

class: center, middle

# Exploring linearity

---

## Data: Paris Paintings

```{r echo=FALSE}
ggplot(data = pp, aes(x = price)) +
  geom_histogram(binwidth = 1000) +
  labs(title = "Prices of paintings")
```

---

## Price vs. width

.question[
Describe the relationship between price and width of painting.
]

```{r echo=FALSE, fig.height=2.75}
ggplot(data = pp, aes(x = Width_in, y = price)) +
  geom_point(alpha = 0.5) +
  labs(x = "Width (in)", y = "Price (livres)")
```

---

### Let's focus on paintings with `Width_in < 100`

```{r}
pp_wt_lt_100 <- pp %>% 
  filter(Width_in < 100)
```

---

## Price vs. width

.question[
Which plot shows a more linear relationship?
]

.small[
  
.pull-left[
```{r fig.width=5, fig.height=4, message=FALSE, echo=FALSE}
ggplot(data = pp_wt_lt_100, 
       mapping = aes(x = Width_in, y = price)) +
  geom_point(alpha = 0.5) +
  labs(title = "Price vs. width", subtitle = "For width < 100 in",
       x = "Width (in)", y = "Price (livres)")
```
]

.pull-right[
```{r fig.width=5, fig.height=4, message=FALSE, echo=FALSE}
ggplot(data = pp_wt_lt_100, 
       mapping = aes(x = Width_in, y = log(price))) +
  geom_point(alpha = 0.5) +
  labs(title = "Log(price) vs. width", subtitle = "For width < 100 in",
       x = "Width (in)", y = "Log(price) (log livres)")
```
]

]

---

## Price vs. width, residuals

.question[
Which plot shows a residuals that are uncorrelated with predicted values from the model?
]

.small[
  
.pull-left[
```{r fig.width=5, fig.height=4, message=FALSE, echo=FALSE}
m_wi_pr <- lm(price ~ Width_in, data = pp_wt_lt_100)
m_wi_pr_tidy <- augment(m_wi_pr)
ggplot(data = m_wi_pr_tidy, 
       mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "red") +
  labs(
    title = "Price vs. width, residuals", 
    subtitle = "For width < 100 in",
    x = "Predicted price (livres)", 
    y = "Residuals"
    )
```
]

.pull-right[
```{r fig.width=5, fig.height=4, message=FALSE, echo=FALSE}
m_log_wi_pr <- lm(log(price) ~ Width_in, data = pp_wt_lt_100)
m_log_wi_pr_tidy <- augment(m_log_wi_pr)
ggplot(data = m_log_wi_pr_tidy, 
       mapping = aes(x = .fitted, y = .resid)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, color = "red") +
  labs(
    title = "Log(Price) vs. width, residuals", 
    subtitle = "For width < 100 in",
    x = "Predicted log(price) (log livres)", 
    y = "Residuals"
    )
```
]

]

--

.question[
What's the unit of residuals?
]

---

## Transforming the data

- We saw that `price` has a right-skewed distribution, and the relationship between price and width of painting is non-linear.

- We also observed signs of the model violation, non-constant variance.

- In these situations a transformation applied to the response variable (y) may be useful.
    - The most common transformation is the log transformation $(\log(y) = ln(y))$ 
    
- This is beyond the scope of the course, but I'm happy to provide guidance if you want to try modeling a response that requires transformation in your final project

---

class: center, middle

## The linear model with multiple predictors

---

### Riders in Florence, MA

The Pioneer Valley Planning Commission collected data in Florence, MA for 90 days from April 5 to November 15, 2005 using a laser sensor, with breaks in the laser beam recording when a rail-trail user passed the data collection station.

- `hightemp`: daily high temperature (in degrees Fahrenheit)
- `volume`:  estimated number of trail users that day (number of breaks recorded)
- `dayType`: weekday or weekend

```{r}
library(mosaicData)
data(RailTrail)
```

---

### Volume vs. Temperature

```{r echo = F}
ggplot(data = RailTrail, aes(x = hightemp, y = volume)) +
  geom_point() + 
  geom_smooth(method = "lm", color = "blue", se = FALSE) +
  labs(x = "High Temperature", y = "Number of Riders", 
       title = "Number of Riders vs. High Temperature") 
```


---

### Volume, Temperature, and Day

.question[
Does the relationship between and `volume` and `hightemp` differ by whether or not it's a weekday or weekend.
]

.small[
```{r echo = F}
ggplot(data = RailTrail, aes(x = hightemp, y = volume, color = dayType)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "High Temperature", y = "Number of Riders", 
       title = "Number of Riders vs. High Temperature", 
       subtitle = "Weekday vs. Weekend", 
       color = "Day Type") 
```
]

---

## Modeling with main effects 

```{r}
m_main <- lm(volume ~ hightemp + dayType, data = RailTrail)
m_main %>% tidy() %>% select(term, estimate) %>%
  kable(format = "markdown", digits =3) #knitr package
```

--

.midi[
Linear model:

$$ \widehat{volume} = -8.747 + 5.348~hightemp + 51.553~ dayTypeweekend $$
]
--

- Plug in 0 for `dayTypeweekend` to get the linear model for weekdays, i.e. days that aren't the weekend.

- Plug in 1 for `dayTypeweekend` to get the linear model for days on the weekend.

---

## Interpretation of main effects

```{r echo = FALSE}
ggplot(data = RailTrail,
       aes(y = volume, x = hightemp, color = dayType)) +
  geom_point(alpha = 0.3) +
  geom_abline(intercept = -8.747, slope = 5.348, color = "#F57670", lwd = 1) +
  geom_abline(intercept = 42.806, slope = 5.348, color = "#1FBEC3", lwd = 1) +
  labs(color = "Day Type")
```

- Weekday (`dayTypeweekend == 0`)
$\widehat{volume} = -8.747 + 5.438~hightemp + 51.553 \times 0$
$= -8.747 + 5.438~hightemp$

- Weekend (`dayTypeweekend == 1`):
$\widehat{volume} = -8.747 + 5.438~hightemp + 51.553 \times 1$
$= 42.806 + 5.438~hightemp$

---

.alert[
- Weekday (`dayTypeweekend == 0`)
$\widehat{volume} = -8.747 + 5.438~hightemp + 51.553 \times 0$
$= -8.747 + 5.438~hightemp$
]

.alert[
- Weekend (`dayTypeweekend == 1`):
$\widehat{volume} = -8.747 + 5.438~hightemp + 51.553 \times 1$
$= 42.806 + 5.438~hightemp$
]

.midi[
- Rate of change in volume as the high temperature increases is the same on days that are weekdays and weekends (same slope).

- Given the same high temperature, days on  the weekend are expected to have higher volume than days that are weekdays (different intercept).
]

---

### Main effects, numerical and categorical predictors

```{r echo=FALSE}
tidy(m_main) %>%
  select(term, estimate) %>% kable(format = "markdown", digits = 3)
m_main_coefs <- tidy(m_main) 
```

.midi[
- For each additional degree Fahrenheit in the day's high temperature, there are predicted to be, on average, `r m_main_coefs %>% filter(term == "hightemp") %>% pull(estimate)` (about 5) additional riders on the trail, holding all else constant.

- Days on the weekend are predicted to have, on average, `r m_main_coefs %>% filter(term == "dayTypeweekend") %>% pull(estimate)` (about 52) more riders on the trail than days that are weekdays, holding all else constant.  

- Weekdays that have a high temperature of 0 degrees Fahrenheit are predicted to have `r m_main_coefs %>% filter(term == "(Intercept)") %>% pull(estimate)` (about -9) riders, on average.
]

---

## What went wrong?

.question[
Why is our linear regression model different from what we got from `geom_smooth(method = "lm")`?
]

.pull-left[
```{r  echo = F}
ggplot(data = RailTrail, aes(x = hightemp, y = volume, color = dayType)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "High Temperature", y = "Number of Riders", 
       title = "Number of Riders vs. High Temperature", 
       subtitle = "Weekday vs. Weekend", 
       color = "Day Type") 
```
]

.pull-right[
```{r echo = FALSE}
ggplot(data = RailTrail,
       aes(y = volume, x = hightemp, color = dayType)) +
  geom_point() +
  geom_abline(intercept = -8.747, slope = 5.348, color = "#F57670", lwd = 1) +
  geom_abline(intercept = 42.806, slope = 5.348, color = "#1FBEC3", lwd = 1) +
  labs(color = "Day Type")
```
]
---

## What went wrong? (cont.)

- The way we specified our model only lets `dayTypeweekend` affect the intercept.

- Model implicitly assumes that days on the weekend and the weekdays have the *same slope* and only allows for *different intercepts*.  

- What seems more appropriate in this case? 
    
    * Same slope and same intercept for both colors
    
    * Same slope and different intercept for both colors
    
    * Different slope and different intercept for both colors?

---

### Interacting explanatory variables

- Including an interaction effect in the model allows for different slopes, i.e. 
nonparallel lines.

- This implies that the regression coefficient for an explanatory variable would 
change as another explanatory variable changes.

- This can be accomplished by adding an <font class="vocab">interaction variable</font> - the product of two 
explanatory variables.

---

### Price vs. hightemp and dayType interaction

.small[
```{r fig.height=2.5}
ggplot(data = RailTrail, aes(x = hightemp, y = volume, color = dayType)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "High Temperature", y = "Number of Riders", 
       title = "Number of Riders vs. High Temperature", 
       subtitle = "Weekday vs. Weekend", 
       color = "Day Type") 
```
]

---

### Modeling with interaction effects

.small[
```{r}
m_int <- lm(volume ~ hightemp + dayType + hightemp*dayType, 
            data = RailTrail)
kable(tidy(m_int) %>% select(term, estimate), format = "html", digits = 3)
```
]

$$\small{\widehat{volume} = \\
-51.224 + 5.980~hightemp + 186.377~dayTypeweekend - 1.906~hightemp \times dayTypeweekend}$$

---

### Interpretation of interaction effects

.vocab[Weekdays]
$$\small{\begin{align}\widehat{volume} &=
-51.224 + 5.980~hightemp + 186.377\times 0 - 1.906~hightemp \times 0\\[5pt]
&= -51.224 + 5.980~hightemp\\\end{align}}$$
<br><br>
--

.vocab[Weekends]

$$\small{\begin{align}\widehat{volume} &=
-51.224 + 5.980~hightemp + 186.377\times 1 - 1.906~hightemp \times 1\\[5pt]
&= -51.224 + 5.980~hightemp + 186.377 - 1.906~hightemp\\[5pt]
&= 135.153 + 4.074~hightemp\end{align}}$$

---

### Interpretation of interaction effects

.alert[
Weekdays:
$$\widehat{volume}-51.224 + 5.980~hightemp$$

Weekends:
$$\widehat{volume} = 135.153 + 4.074~hightemp$$
]

.midi[
- Rate of change in volume as the high temperature increases is different on days that are weekdays versus those that are weekends (**different slope**)

- Given the same high temperature, days on the weekend are expected to have higher volume than days that are weekdays (**different intercept**).
]

---

### Interpretation of interaction effects

- Weekdays: $\widehat{volume}-51.224 + 5.980~hightemp$

- Weekends: $\widehat{volume} = 135.153 + 4.074~hightemp$


```{r fig.height=3, echo = FALSE}
ggplot(data = RailTrail, aes(x = hightemp, y = volume, color = dayType)) +
  geom_point() + 
  geom_smooth(method = "lm", se = FALSE) +
  labs(x = "High Temperature", y = "Number of Riders", 
       title = "Number of Riders vs. High Temperature", 
       subtitle = "Weekday vs. Weekend", 
       color = "Day Type") 
```

---

## Third order interactions

- Can you? Yes

- Should you? Probably not if you want to interpret these interactions in context of the data.

---

### Practice

Suppose you wish to fit a model using `hightemp` and `summer` to predict the number of riders on a trail. 

- `summer`: 1 if the day is during the summer, 0 otherwise

```{r echo = F}
RailTrail <- RailTrail %>% mutate(summer = as.factor(summer))
m_int2 <- lm(volume ~ hightemp + summer + hightemp*summer, data = RailTrail)
kable(tidy(m_int2) %>% select(term, estimate), format = "html", digits = 3)
```

.question[
1. Interpret the coefficient of `summer1`. 
2. Interpret the intercept. Is this interpretation meaningful?
2. Write the model equation for days that are <u>not</u> during the summer. 
3. Write the model equation for days that are during the summer.
]

---

class: middle, center

## Exam 01 

---

### Exam 01 

- Exam grades will be released after class

- Code for the exam may be found in the **Resources** folder on [Sakai](http://sakai.duke.edu)

- Review the code solutions and feedback in Gradescope! 
    - Attend office hours if you have any questions about your exam grading. 
    
- Be careful about joins! 
    - Only use a join when needed!
    - Joins can be computationally intensive, so make sure you're using the correct one!
    
---

### Writing better code

- We want to make sure the code we right is not only "correct" but also efficient

- This means 
    - Not unnecessarily saving output
    - Not unnecessarily repeating processes
    - Using the simplest code possible!

- Writing code is an iterative process!
    - The first draft isn't always the best draft!
    
---

### Practice: Writing better code

- Copy the **Writing better code** project in RStudio Cloud

- The file contains two pieces of code that perform the correct calculations but needs revision!

- Rewrite the code so that it performs the same calculations but does so in a simpler and less computationally intensive way. 



    