---
title: "Math 218: Statistical Learning"
author: "Survival Analysis"
date: "9/14/2022"
output: 
   xaringan::moon_reader:
    css: "math218-slides.css"
    nature: 
      highlightLines: true
      highlightStyle: github
      countIncrementalSlides: false
      slideNumberFormat: "%current%"  
editor_options: 
  chunk_output_type: console
---


```{r}
knitr::opts_chunk$set(echo = F, warning = F,
                      message = F, fig.align = "center", fig.width = 5, fig.height = 5)
library(tidyverse)
library(ISLR2)
library(survival)
library(DT)
library(ggfortify)

```

class: center, middle

# Housekeeping

---

## Survival Analysis

- Survival analysis concerns a special kind of outcome/response variable: the **time until an event occurs**

  - Suppose we are conducting a three-year medical study about patients who have had one heart attack
  
  - We would like to fit a model to predict the time until the second heart attack, using features such as baseline health measurement, weight, sex, etc.
  
--

- May sound like a regression problem, but important complication: survival data is often **censored**

---

## Survival and Censored data

- **Censoring** refers to a type of missing data

- Why? It is very likely that some of the patients:

  - drop out of the medical study, or
  
  - do not have a second heart attack within the three-year study window
  
--

- Does not mean the patient will never experience the event, but that the event *did not happen while under observation*

--

- Do not want to discard this subset of patients, because surviving beyond study window is still valubale information! 

---

## Another example

- A company (e.g. Netflix) wants to model *churn*, the event when customers cancel their subscription to a service (e.g. account)

- Netflix might collect data on customers over some timer period in order to predict each customer's time to cancellation

  - Presumably, not all customers will have cancelled their subscription by the end of this time period
  
  - Time to cancellation is censored for this group of customers
  
---

## Survival and Censoring Times

- For each individual, we suppose there is a true

  - **survival**/**failure**/**event** time $T$, and
  
  - **censoring** time $C$
  
- Survival time $T$ represents the time at which the event of interest occurs (e.g. heart attack)

- Censoring time $C$ is time at which the censoring occurs (e.g. study ends or patient drops out)

---

## What we observe

- We observe *either* the survival time $T$ or the censoring time $C$ (not both)

- Specifically, we observe the random variable $$Y = \min(T,C)$$

--

- If the event occurs before censoring, do we observe $T$ or $C$?

- If the censoring occurs before event, do we observe $T$ or $C$?

--

- Thus, we also observe status indicator:

$$\delta = \begin{cases}1 & \text{ if } T \leq C  \\ 
0 & \text{ if } T > C\end{cases}$$

--

- In our data, we observe $n$ pairs $(y_{i}, \delta_{i}), \ldots, (y_{n}, \delta_{n})$

---

## Illustration

---


## Closer look

- Suppose we have data from a study on cancer treatment over a five-year window. Want to model time until death

- Many patients drop out of the study early because they are very sick

  - Our analysis should incorporate the reason why the patients dropped out early. Why?

- Suppose that males who are very sick are more likely to drop out than females who are very sick

  - Comparison of male and female survival times may wrongly suggest that males survive longer
  
--

- **Assumption**: conditional on the features $X$, the event time $T$ is *independent* of the censoring time $C$

---

## Types of censoring

- Left, right, and interval

- We will focus on *right*-censoring: when $T \geq Y$ (i.e. true event time is at least as large as observed time)

---


## Survival Curve

- **Survival curve** or function is defined as

$$S(t) = \text{Pr}(T > t)$$

  - i.e. $S(t)$ is the probability that we survive beyond time $t$
  
--

- $S(t)$ is a decreasing function of $t$. Why?

--

- For given $t$, a larger $S(t)$ implies lower probability that patient will die before time $t$

---

## Brain Cancer Data

```{r}
datatable(BrainCancer, options = list(pageLength = 5)) %>%
  formatStyle(columns = 1:ncol(BrainCancer), fontSize = '90%')
```

---

## Brain Cancer Data

- Survival times for patients with primary brain tumors undergoing treatment with stereotactic radiation methods

- $\color{blue}{\text{sex}}$: Female, Male

- $\color{blue}{\text{diagnosis}}$: Meningioma, LG glioma, HG glioma, Other

- $\color{blue}{\text{loc}}$ (location): Infratentorial, Supratentorial

- $\color{blue}{\text{ki}}$: Karnofsky index

- $\color{blue}{\text{gtc}}$: Gross tumor volume (cm$^3$)

- $\color{blue}{\text{stereo}}$ (method): SRS, SRT

- $\color{orange}{\text{status}}$ (whether still alive at end of study): 0 = Yes, 1 = No

- $\color{orange}{\text{time}}$:  the time to the event (death or censoring), in months

---

## Brain Cancer Data

How long was the study?

How many patients still alive at end of study?

```{r echo = T}
max(BrainCancer$time)

table(BrainCancer$status)
```

---


```{r}
BrainCancer %>%
  mutate(status = as.factor(status)) %>%
  ggplot(., aes(x = time, y = status)) +
  geom_boxplot() +
  ggtitle("Time until event", subtitle = "0 = still alive, 1 = died")
```

---


## Estimating Survival Curve

- Suppose we want to estimate $S(20) = \text{Pr}(T > 20)$, the probability that a patients survives for at least 20 months. 

question[How might we do that?]

---

## Estimating Survival Curve

- Should we compute the proportion of patients who are known to have survived beyond 20 months (i.e. $Y > 20$)? 

```{r echo = T}
n <- nrow(BrainCancer)
(n_surv20 <- sum(BrainCancer$time > 20))
n_surv20/n
```

- Issue: those with $y_{i} < 20$ did not necessarily die!

```{r echo = T}
# number of patients who dropped-out before 20 months
sum(BrainCancer$time < 20 & BrainCancer$status == 0)
```

---

## Estimating Survival Curve (cont.)

- Instead, maybe we compute the proportion of patients for whom $Y>20$ among the patient who were *not* censored by time $t = 20$

```{r}
n_not_cens20 <- sum(BrainCancer$time < 20 & BrainCancer$status == 0)
n_surv20/(n - n_not_cens20)
```

- Issue: ignores patients who were censored before $t = 20$

---

### Kaplan-Meier estimator of survival curve

```{r fig.width=9, fig.height=5}
attach(BrainCancer)
fit.surv <- survfit(Surv(time, status) ~ 1)
autoplot(fit.surv) +
  labs(x = "Time (months)", y = "Estimated probability of survival") +
  ggtitle("Kaplan-Meier Estimate of Survival Curve for Brain Cancer data")+
  theme(text = element_text(size = 16))
```

--

- Based on KM estimate, approximately what is $S(20)$?

  - How does this compare to our "naive" estimates?
  
---

### Stratifying

```{r fig.width=9, fig.height=5}
fit.surv.strat <- survfit(Surv(time, status) ~ sex)
autoplot(fit.surv.strat, conf.int = F) +
  labs(x = "Time (months)", y = "Estimated probability of survival") +
  ggtitle("Kaplan-Meier Estimate of Survival Curve for Brain Cancer data",
          subtitle = "Stratified by sex")+
  theme(text = element_text(size = 16)) +
  labs(col = "sex")
```

---

## Log-rank test?

---

## Regression with survival response

- Want to learn how features $X$ might influence the probability of survival. 

- How to fit a regression model to survival data? 

--

- Want to predict true survival time $T$, but we only observe $Y = \min(T, C)$

- Solution: sequential construction

---

## Hazard function

- The **hazard function** is formally defined as

$$h(t) = \lim_{\Delta t \rightarrow 0} \frac{\text{Pr}(t < T \leq t + \Delta t| T > t)}{\Delta t}$$
where $T$ is true survival time.

- Interpretation: death rate in the instant right after time $t$, given survival up to time $t$

--

- Implies that for arbitrarily small $\Delta t$,

$$h(t) \approx \frac{\text{Pr}(t < T \leq t + \Delta t| T > t)}{\Delta t}$$
---

## Hazard function and Survival curve

$$
\begin{align*}
h(t) &= \lim_{\Delta t \rightarrow 0}\frac{\text{Pr}(t < T \leq t + \Delta t| T > t)}{\Delta t} \\
&= \lim_{\Delta t \rightarrow 0}\frac{\text{Pr}((t < T \leq t + \Delta t) \cap (T > t))/\Delta t}{\text{Pr}(T > t)} \\
&= \lim_{\Delta t \rightarrow 0}\frac{\text{Pr}(t < T \leq t + \Delta t)/\Delta t}{S(t)} \\
&= \frac{f(t)}{S(t)}
\end{align*} 
$$

where $f(t) = \lim_{\Delta t \rightarrow 0}\frac{\text{Pr}(t < T \leq t + \Delta t)}{\Delta t}$ is probability density function of $T$

---

## Proportional Hazards Model

- We would like to model survival time as a function of predictors. 

- There are many ways, but we will discuss the **proportional hazards** model which assumes

$$h(t | x_{i}) = h_{0}(t) \exp\left(\sum_{j=1}^{p} x_{ij} \beta_{j}\right)$$

- $h_{0}(t) \geq 0$ is the *baseline hazard*, typically unspecified function

  - Why baseline? Hazard function for individual with $x_{i1} = \ldots = x_{ip} = 0$

- Name "proportional" hazards comes from the fact that the hazard function given a set of predictors $x_{i}$ is simply proportional to the baseline

  - The factor $\exp\left(\sum_{j=1}^{p} x_{ij} \beta_{j}\right)$ is called the **relative risk** for feature vector $\mathbf{x}_{i} = (x_{i1}, \ldots, x_{ip})'$, relative to that for $\mathbf{x}_{0} = (0, \ldots, 0)'$
  
---

## Proportional Hazards Model

- What do we mean that $h_{0}(t)$ is unspecified?

  - We make no assumptions about its functional form
  
  - Allows the instantaneous probability of event (e.g. death) at time $t$ to take any form (given survival to time $t$)
  
  - Extremely flexible, allowing a wide range of relationships between $x$'s and the survival time
  
--

- Our only assumption: a one-unit increase in $x_{ij}$ corresponds to an increase in $h(t|x_{i})$ by a *factor* of $e^{\beta_{j}}$

--

.question[Is proportional hazards model parametric or nonparametric?]

---

## Example: proportional hazards

- $p=1$ where $x$ is binary predictor (blue, red). Plots of $\log h(t)$ and $S(t)$

```{r prop_hazard, fig.width=10, fig.height=5}
time <- seq(0.05,3,0.1)
h0 <- function(t){
  (t/2)^1.5
}

b <- 1

h0_t <- c(h0(time)*exp(b*1) , h0(time)*exp(b*0))
df1 <- data.frame(x = c(rep(c(1,0), each = length(time))),
                  t = rep(time,2),
                  stat = "Log Hazard",
                  val= log(h0_t)) 

S1 <-  S0 <- rep(NA, length(time))
for(t in 1:length(time)){
  S1[t] <- exp(-exp(b*1)*integrate(h0, lower = 0, upper = time[t])$value)
  S0[t] <- exp(-exp(b*0)*integrate(h0, lower = 0, upper = time[t])$value)
}

df2 <- data.frame(x = c(rep(c(1,0), each = length(time))),
                  t = rep(time,2),
                  stat = "Survival Probability",
                  val = c(S1, S0)) 

rbind(df1,df2) %>%
  mutate(x = factor(x)) %>%
  ggplot(., aes(x = t, y = val, col = x))+
  geom_line()+
  facet_wrap(~stat, scales = "free_y")+
  theme(text = element_text(size = 16)) +
  guides(col = "none") +
  labs(x = "Time", y = "")

```

---

## Example: non-proportional hazards

```{r non_prop_hazard, fig.width=10, fig.height=5}
h0_t <- c(h0(time)*exp(b*1) , h0(time)*exp(b*0)+ 0.5)
df1 <- data.frame(x = c(rep(c(1,0), each = length(time))),
                  t = rep(time,2),
                  stat = "Log Hazard",
                  val= log(h0_t)) 

S1 <-  S0 <- rep(NA, length(time))
for(t in 1:length(time)){
  S1[t] <- exp(-exp(b*1)*integrate(h0, lower = 0, upper = time[t])$value)
  S0[t] <- exp(-(exp(b*0)*integrate(h0, lower = 0, upper = time[t])$value + time[t]*0.5))
}

df2 <- data.frame(x = c(rep(c(1,0), each = length(time))),
                  t = rep(time,2),
                  stat = "Survival Probability",
                  val = c(S1, S0)) 

rbind(df1,df2) %>%
  mutate(x = factor(x)) %>%
  ggplot(., aes(x = t, y = val, col = x))+
  geom_line()+
  facet_wrap(~stat, scales = "free_y")+
  theme(text = element_text(size = 16)) +
  guides(col = "none") +
  labs(x = "Time", y = "")

```

---

## Estimating coefficients

- How to estimate the $\beta_{j}$'s?

- Because $h_{0}(t)$ is unknown, cannot straightforwardly estimate the $\beta_{j}$'s as usual

--

- However, we  can still estimate $\beta$ *without having the specific form of* $h_{0}(t)$ using the partial likelihood

---

## Partial likelihood

- Assume that the failure/event times are unique (i.e. only one patient dies at a particular time), and $i$-th observation is uncensored

  - Thus, observation $y_{i}$ is the failure time
  
- Hazard function for $i$-th observation at time $y_{i}$ is $h(y_{i}|x_{i}) = h_{0}(y_{i}) \exp\left(\sum_{j=1}^{p} x_{ij}\beta_{j}\right)$

- Total hazard at time $y_{i}$ for the *at risk* observations is

$$\sum_{i': y_{i'} \geq y_{i}} h_{0}(y_{i})\exp\left(\sum_{j=1}^{p} x_{i'j}\beta_{j}\right)$$
---


## Partial likelihood (cont.)

- Thus, probability that the $i$-th observation is the one to fail at time $y_{i}$ is

$$\frac{h_{0}(y_{i}) \exp\left(\sum_{j=1}^{p} x_{ij}\beta_{j}\right)}{\sum_{i': y_{i'} \geq y_{i}} h_{0}(y_{i})\exp\left(\sum_{j=1}^{p} x_{i'j}\beta_{j}\right)} = \frac{ \exp\left(\sum_{j=1}^{p} x_{ij}\beta_{j}\right)}{\sum_{i': y_{i'} \geq y_{i}} \exp\left(\sum_{j=1}^{p} x_{i'j}\beta_{j}\right)}$$

- The baseline function is cancelled out!

--

- Partial likelihood (PL): product of these probabilities over all the uncensored observations:

$$\text{PL}(\beta) = \prod_{i:\delta_{i} = 1}\frac{ \exp\left(\sum_{j=1}^{p} x_{ij}\beta_{j}\right)}{\sum_{i': y_{i'} \geq y_{i}} \exp\left(\sum_{j=1}^{p} x_{i'j}\beta_{j}\right)}$$

  - Use PL to estimate the $\beta_{j}$'s, as it is valid regardless of true $h_{0}(t)$!
  
---

## Illustration

---

## Proportional Hazards (details)

- There is no intercept $\beta_{0}$

- This presentation assumed that there are no ties in failure times (i.e. each $y_{i}$ for uncensored $i$ is unique)

- The baseline $h_{0}(t)$ can be estimated, but is beyond the scope here

---

## Brain Cancer data

```{r}
fit.cox <- coxph(Surv(time, status) ~ ., data = BrainCancer)
coeff_sum <- (summary(fit.cox)$coefficients)[,c(1,3,4,5)]
round(coeff_sum, 3)
```

- Estimated hazard for a male patient is $e^{0.184}$ = 1.202 times greater than for a female patient

  - Holding other features fixed, males have a 1.2 times greater chance of dying than females, at any point in time. But not significant
  
--

- Each one-unit increase in the Karnofsky index corresponds to a multiplier of $e^{-0.055}$ = 0.946 in the instantaneous chance of dying

  - The higher the Karnofsky index, the lower the chance of dying at any given point in time

