---
title: "Bootstrap Implementation"
author: ""
output: html_document
date: "10/7/2022"
---

```{r include = F}
library(tidyverse)
```

## The Bootstrap

The Bootstrap can be using in almost any scenario, which makes it extremely appealing. One area where this is especially useful is learning about the distribution of a statistic. Suppose we have observed some data `z` and we don't know the underlying distribution. We want to learn about the distribution of the *median*. 


Here is some data:

```{r generate_data}
set.seed(2)
n <- 20
z <- rgamma(n, 5, 2)
```

```{r echo = F, fig.align="center", fig.height=4, fig.width=4}
ggplot(data.frame(z = z), aes(x=z)) +
  geom_histogram(bins = 15) +
  geom_vline(xintercept = median(z), col = "blue", linetype = "dashed") +
  ggtitle(paste0("Observed median: ", round(median(z),2)))
```

Our best guess at the median $m$ is probably the median of the observed data. But how confident are we? We can bootstrap to get a range of reasonable/plausible values for the median of this data.

### Part 1

Write code that generate bootstrap samples for the median $m$. Don't forget to set a seed for reproducibility!

```{r bootstrap}
## your code here

```


### Part 2

Now you can calculate the standard error of your estimate for median of `z` using the `sd()` function on your vector!

**What is the standard error of your estimate?**

```{r bootstrap_se}
## your code here
```


### Part 3

Now plot the distribution of your bootstrap medians using a histogram. Add vertical lines for the 95% bootstrap intervals (i.e. one at the 2.5-th percentile, and one at the 97.5-th percentile). The quantiles can be accessed using the `quantile()` function.


**You should clearly understand how this plot differs from the one above.**
