---
title: "Lab 07: Two-sample inference"
subtitle: "due Tue, June 29 at 11:59p"
output: 
  tufte::tufte_html:
    highlight: pygments
    css: "sta199-labs.css"
link-citations: yes
---

```{r include=FALSE}
library(tidyverse)
library(tufte)
library(knitr)
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  digits = 2
  )
knitr::opts_chunk$set(eval = FALSE)
```


# What makes a good wine?

The goal of today's lab is to practice two-sample statistical inference using
both simulation- and CLT-based approaches. Use the lecture notes, readings, 
and application exercises to help you complete the lab.

# The Data

Today's dataset has been adapted from a large data set of wine reviews from
[Kaggle](https://www.kaggle.com/zynicide/wine-reviews). Reviews were scraped from 
WineEnthusiast, which contain information about both the wine itself as well as 
the reviewer's description of that wine. Here, I provide a subset of the data. 
You may consider each of these observations to be an
independent, representative sample of all wines.

The variables are as follows

- `country`: the country that the wine is from (France, Italy, US)
- `points`: the number of points WineEnthusiast rated the wine on a scale of 1-100. 
- `price`: cost for a bottle (USD)
- `variety`: type of grapes used to make the wine
- `type`: red or white wine
- `title`: title of the wine review
- `description`: reviewer's description of the wine

The data for today's lab may be found by cloning your repository available at 
the class GitHub repository. Load the data into your RStudio environment, and
don't forget to configure GitHub beforehand. Write all R code according to the 
style guidelines discussed in class. 


To following resource provides code needed to make useful symbols. You may use
the code to typeset the characters of interest in the *narrative* of your
document:

- $\mu$: `$\mu$`
- $\alpha$: `$\alpha$`
- $\ge$: `$\ge$`
- $\le$: `$\le$`
- $\neq$: `$\neq$`
- $H_0$: `$H_0$`
- $H_a$: `$H_a$`


# Configure git

```{r eval=FALSE}
library(usethis)
use_git_config(user.name = "github username", user.email = "your email")
```



# Exercises

**Overall hint:** When performing a hypothesis test, you must provide the 
significance level of your test, the null and alternative hypotheses, the 
p-value, your decision, and an interpretation of the p-value in context of the 
original research question. 

You may load in the data with the following code, where `____` should be
replaced by a meaningful name of your choosing:

```{r eval=FALSE}
library(tidyverse)
library(infer)
____ <- read_csv("data/wines.csv")
```

**At the start of each exercise that requires simulation, set a random seed**
**equal to the exercise number in the R chunk.**

1. Suppose you are interested in knowing if red or white wines in these three countries
are more rated heigher.
Comprehensively evaluate the hypothesis that the average `points` of
white wines is lower than the average `points` of red wines using a simulation-based approach.


2. Now suppose you are now interested in how the type of wine (red or white) impacts
its price. Comprehensively evaluate the hypothesis that
the average price of red wines is different from the average price of white wines
in these three countries. Use a CLT-based approach at the $\alpha = 0.05$ level.


3. Construct and interpret a a 95% two-sided confidence interval for the 
difference you investigated in Exercise 2 using a CLT-based approach.


4. For this exercise, consider wines from US and Italy only. Evaluate the hypothesis
that the `country` where the wine is produced is independent of its `type` (red or white).
Use a simulation-based approach at the $\alpha = 0.05$ level. Visualize your null distribution.


5. Let's say I'm doing to a nice dinner party, and I'm told to bring a bottle of wine. 
I want to bring an acceptable wine, but I don't like these friends enough to spend more 
than $20 on a bottle of wine, so I have to be strategic. I've been told that a cheap white
is usually better than a cheap red, but that an expensive red is always better than an expensive white. Maybe I should buy a wine based on its points-to-price ratio. 

Create a new variable for this ratio. 
Wines with higher values for this ratio might indicate that the wine is a good deal. 

Then, using a CLT-based approach, comprehensively evaluate the hypothesis at the $\alpha = 0.05$ level
that among wines that are at most $20 per bottle, 
white wines have a higher average point-to-price ratio compared to red wines. 

How would your decision and conclusion change if we used $\alpha = 0.10$ instead?


```{marginfigure}
**Hint**: refer back to the lectures about probability and the definition of Type I error
to answer this question! 
```

6. Suppose you perform 10 independent hypothesis tests at the $\alpha = 0.05$
level, and further suppose that in reality, the null hypothesis is TRUE for all
10 tests. What is the probability that you make at least one Type I error?


7. Let's say you just really hate white wine and are trying to demonstrate to
others that it's associated with lower ratings in terms of points. Describe the
potential ethical ramifications of performing all of these tests and reporting
only the significant results from the tests that support your narrative. 
Consider your answer from Exercise 6 in crafting this narrative.

```{marginfigure}
**Hint**: if your distribution is skewed, then median and IQR are appropriate 
summary statistics. If your distribution is symmetric, then mean and standard
deviation are appropriate.
```

8. Plot and describe the distribution of the `points` of wine. Also provide 
appropriate summary statistics when describing your distribution, obtained by
using the `summarise()` function.


9. From Ex. 4, you should have noticed that the values that `points` takes on lie
between 80 and 100, even though the points are awarded on the 1-100 scale. 
This is because WineEnthusiast only provides reviews for wines that are at least 
considered "Good". They using the following scale for classification of quality:

  - 80-84 = Good
  - 85-89 = Very good
  - 90-94 = Outstanding
  - 95-100 = Classic
  
Create a new variable for the quality of each wine based on its points.

Is there evidence to suggest that the quality of a wine and the `country` where its produced
are associated?
Comprehensively evaluate this hypothesis using an appropriate method of your choice.


# Submission

Knit to PDF to create a PDF document. Knit and commit all remaining changes, 
and push your work to GitHub. Make sure all files are updated on your GitHub 
repo.

Please only upload your PDF document to Gradescope. Associate the "Overall"
graded section with the first page of your PDF, and mark where each answer is 
to the exercises. If any answer spans multiple pages, then mark all pages.

