---
title: "Lab 05 - Bootstrap estimation"
subtitle: "due Thursday, June 10 at 11:59p"
output: 
  tufte::tufte_html:
    highlight: pygments
    css: "sta199-labs.css"
    toc: true
    toc_depth: 2
link-citations: yes
---

```{r include=FALSE}
library(tidyverse)
library(tufte)
library(knitr)
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  digits = 2
  )
knitr::opts_chunk$set(eval = FALSE)
```

# Introduction and Data

The goal of today's lab is to practice creating bootstrap confidence intervals,
and visualizing bootstrap distributions. 

# Tips 

- Use a small number of reps (about 100) as you write and test out your code. Once you have finalized all of your code, increase the number of reps to at least 1000 (10,000 if possible) to produce your final results. 
- Write your code and narrative in a reproducible way, so you can easily change the number of reps. Tips on making your code more reproducible:

### Tip 1: Set your seed

When dealing with randomness (as often the case in simulation in statistics), it is important to specify which pseudo-random draw you used in your analysis, so that you or someone else can reproduce the exact numbers you initially report. The `set.seed()` function in R allows you to ensure that all of your analysis relies on a specific pseudo-random draw:

```{r set_seed, eval = F}
set.seed(42)
```

### Tip 2: Define global parameters

Often, we rely on specific parameters values throughout our analysis, and at a later point, we may want to replace them. In order to minimize the need to change your code later, we can assign the parameter values to a name, and use the name (rather than the hard-coded value) downstream. Then, to update your code at a later point, you can just change the value. Here, we are assinging the number of reps 
to a variable called `num_reps`:

```{r global_params, eval = F}
num_reps <- 100
boot_dist <- diamonds %>%
  infer::specify(response = price) %>%
  infer::generate(reps=num_reps, type = "bootstrap") %>%
  infer::calculate(stat="mean")
```

# The data

The data for today's lab may be found by cloning your **lab-05-durham-**  repository available in the [GitHub course organization](https://github.com/sta199-summer2021).

Today's data comes from the City of Durham's annual Resident Satisfaction 
Survey for 2018 (more information accessible 
[**here**](https://www.dconc.gov/government/departments-a-e/county-manager/city-county-resident-survey)). In particular, the `durham_survey` dataset contains 
data from 608 Durham residents on the survey questions below. Assume that the
data are representative of Durham residents and may be generalized to the wider
population of all city residents.

Any variable starting with `quality` refers to the perceived quality of the
listed variable, with 1 being "highly dissatisfied," 3 being "neutral", and
5 being "highly satisfied." A value of 9 indicates that the subject responded
with N/A.

You may load in the data with the following code, where `____` should be
replaced by a meaningful name of your choosing:

```{r eval=FALSE}
library(tidyverse)
library(infer)
____ <- read_csv("data/durham_survey.csv")
```

# Exercises

Write all R code according to the style guidelines discussed in class. 

```{marginfigure}
**Hint**: be careful with how missing values are coded in this survey. As well,
don't forget to set a seed in order to ensure reproducibility!
```

1. Provide a point estimate of the mean satisfaction with the fire department 
(`quality_fire`) among Durham residents in 2018.

2. Construct a 95%  bootstrap interval for the mean satisfaction with
the fire department among durham residents in 2018. Use at least 1,000 bootstrap
samples. Make sure your interval is *reproducible*.

3. Visualize the bootstrap distribution and your confidence interval from 
Exercise 2. Interpret the confidence interval you constructed.

4. Provide a point estimate of the *proportion* of the respondents in the 
survey who were satisfied (score of 4 or 5) with the quality of parks and
recreation (`quality_parks_rec`) in Durham. 

```{marginfigure}
**Hint**: see if you can reuse parts of code used in previous exercises.
```

5. Construct a **99%**  bootstrap interval for the proportion of 
respondents in the survey who were satisfied with the quality of parks and 
recreation in Durham. Make sure your interval is *reproducible*.

6. Visualize the bootstrap distribution and your confidence interval from 
Exercise 5. Interpret the confidence interval you constructed.

```{marginfigure}
**Hint**: If *either* of the two scores is missing, then that observation cannot 
be used to calculate the correlation.
```

7. Provide a point estimate of the *correlation* between survey scores of
perceived quality of bike paths (`quality_bike_path`) and the perceived quality 
of pedestrian paths (`quality_ped_path`). 

```{marginfigure}
**Hint**: To simulate the correlation between two variables, use `specify(var1 ~ var2)`. Remember that correlation is still a numerical quantity, so that should help you choose the type of simulation you want to perform.
```

8. Construct a 95% bootstrap interval for the correlation between
survey responses for perceived quality of bike paths and pedestrian paths. Make sure your interval is *reproducible*.


9. Construct a **99%** bootstrap interval using the bootstrap distribution from Exercise 8.  

    How does the 99% bootstrap interval compare to the 95% bootstrap interval calculated in the previous exercise?
  
    In general, how does the bootstrap interval change when the confidence level increases?

# Submission

Knit to PDF to create a PDF document. Knit and commit all remaining changes, 
and push your work to GitHub. Make sure all files are updated on your GitHub 
repo.

Please only upload your PDF document to Gradescope. Associate the "Overall"
graded section with the first page of your PDF, and mark where each answer is 
to the exercises. If any answer spans multiple pages, then mark all pages.

