---
title: "Lab 04 - Resampling"
subtitle: "Due Sunday, 10/9 at 11:59pm"
output: 
  tufte::tufte_html:
    css: "./math218-labs.css"
    tufte_variant: "envisioned"
    highlight: tango
    toc: true
    toc_depth: 1
editor_options: 
  chunk_output_type: console
---

```{r message = F, warning = F}
library(tidyverse)
```

# Introduction

This lab explores the resampling techniques learned in class this week (cross-validation and the bootstrap). You will most likely need to install the package `boot`. The following code generates some data for us to use.

```{r}
# generate data
set.seed(1)
n <- 150
x <- rnorm(n)
eps <- rnorm(n)
y <- -1 + 2*x + 0.5*x^2 + eps
df <- data.frame(x,y)
```

# Leave-one-out CV (LOOCV)


The LOOCV estimate can be automatically computed for any generalized linear model using the `glm()` and `cv.glm()` functions in `R`. The latter comes from the `boot` library. 

First we fit our model to the data

```{marginfigure}
Note: the default `family` for `glm()` is gaussian, but we are specifying it here just to be safe!
```

```{r}
glm.fit <- glm(y~x, family = gaussian, data= data.frame(y=y,x=x))
```

Then we can use the `cv.glm()` function to perform cross-validation. We pass in, at minimum, the `data` and the fitted model `glmfit`. We could also specify the number of groups $K$, but the default is $K = n$ (i.e. LOOCV).

```{marginfigure}
Recall that there is not random splitting in LOOCV, so you should get the same LOOCV error every time.
```


```{r}
library(boot)
cv.err <- cv.glm(data = data.frame(y=y,x=x), glmfit = glm.fit)
```

The LOOCV error estimate can be accessed as follows:

```{r}
cv.err$delta[1]
```

# $k$-fold CV

As alluded to above, the `cv.glm()` function can also be used to implement $k$-fold CV. A common choice for $k$ is 5 or 10. Because $k$-fold randomly splits the data, it's good practice to set a seed to ensure the same results every time.

```{r}
set.seed(10)
cv.err10 <- cv.glm(data = data.frame(y=y,x=x), glmfit = glm.fit, K = 10)
cv.err10$delta[1]
```

We see that the LOOCV and 10-fold CV estimates are very similar. 



# The Bootstrap

The Bootstrap can be using in almost any scenario, which makes it extremely appealing. One area where this is especially is learning about the distribution of a statistic. Suppose we have observed some data `x` and we don't know the underlying distribution. We want to learn about the distribution of the *median*. 

Suppose we have observed some (generated) data `x` and we don't know the underlying distribution. We want to learn about the distribution of the *median*. 

```{r}
set.seed(2)
n <- 20
x <- rgamma(n, 5, 2)
```


The `boot()` function automates the bootstrap approach. You pass in three arguments: 1) the `data`, 2) a function that calculates the `statistic` (quantity), and 3) the number of repetitions `R`.

Because I am interested in learning about the median, I will write a function `med_fun()` that calculates the median to pass into the `statistic` argument. The function must take at least two arguments: 1) the data, and 2) a vector of indices that define the bootstrap sample. In the following code, these arguments are `x` and `inds`, respectively. 

```{r}
med_fun <- function(x, inds){
  median(x[inds])
}
```

Then I will run `R = 1000` bootstrap samples on the collected `x` data to calculate the median using my new `med_fun` function.

```{r}
set.seed(1)
boot_out <- boot(data = x, statistic = med_fun, R = 1000)
```

The bootstrap estimate for the standard error of the median is $\text{SE}(\hat{m}) = $`r round(sd(boot_out$t), 3)`.


# YOUR TURN!

## Exercise 1: Bootstrap practice

We will revisit the `bmd` data that we saw last week. We will predict the probability of a `fracture` status given the patient's bone mineral density `bmd` and height `height_cm`.

In particular, we will now compute estimates for the standard errors of the logistic regression coefficients in two different ways: (1) using the bootstrap, and (2) using the standard formula for computing the standard errors in the `glm()` function. Do not forget to set a random seed before beginning your analysis.


```{r}
bmd_dat <- read.csv("data/bmd.csv", header = T) %>%
  mutate(status = factor(fracture, levels = c("no fracture", "fracture")))
```

a) Using the `summary()` and glm() `functions`, determine the estimated standard errors for the coefficients associated with `bmd` and `height_cm` in a multiple logistic regression model that uses both predictors.


b) Write a function that takes in as input the `bmd_dat` dataset as well as an index of the observations, and outputs the coefficient estimates for `bmd` and `height_cm` in the multiple logistic regression model.

c) Use the `boot()` function as well as your written function in (c) to estimate the standard errors of the logistic regression coefficients. Don't forget to set a seed.

d) Comment on the estimated standard errors obtained using the `glm()` function and using your bootstrap function.


## Exercise 2: More bootstrap practice

We will work with `Boston` dataset from the `ISLR2` package, which contains housing values in 506 suburbs of Boston.

```{r}
library(ISLR2)
data("Boston")
```

a) Based on this data set, provide an estimate for the population mean of the median value of owner-opccupied homes `medv` (in thousands). Call this estimate $\hat{\mu}$.


b) Provide an estimate of the standard error of $\hat{\mu}$. Interpret this result. *Hint: We can compute the standard error of the sample mean by dividing the sample standard deviation by the square root of the number of observations.*

c) Now estimate the standard error of $\hat{\mu}$ using the bootstrap. How does this compare to your answer from (b)?


d) Based on your bootstrap estimate from (c), provide a 95 % confidence interval for the mean of `medv`. Compare it to the results obtained using `t.test(Boston$medv)`.
*Hint: You can approximate a 95 % confidence interval using the formula* $\hat{\mu} \pm 2 \text{SE}(\hat{\mu})$


```{marginfigure}
Hint: use the `quantile()` function
```

e)  Based on this data set, provide an estimate for the fifteenth percentile of `medv` in Boston census tracts. Call this quantity $\hat{p}_{15}$.


```{marginfigure}
Try writing a function that allows you to pass in the percentile, rather than hard coding the fifteenth percentile.
```

f) Use the bootstrap to estimate the standard error of $\hat{p}_{15}$. This will entail writing a function to calculate the 15-th percentile. Comment on your findings. 

