---
title: "Lab 04: Examining smoking and health outcomes"
subtitle: "due Tue, June 8 at 11:59p"
output: 
  tufte::tufte_html:
    tufte_variant: "envisioned"
    highlight: pygments
    css: "./sta199-labs.css"
    toc: true
    toc_depth: 1
link-citations: yes
---

```{r include=FALSE}
library(tidyverse)
library(tufte)
library(knitr)
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  digits = 2
  )
knitr::opts_chunk$set(eval = FALSE)
```

# Introduction 

The goal of today's lab is to practice visualizing and calculating probabilities using the tidyverse. 

# Set-up
Clone the **lab-04-whickham** assignment repo in RStudio. Don't forget to configure git if you haven't already done so:

```{r eval = FALSE}
library(usethis)
use_git_config(user.name="your github username", user.email="your email")
```

# Packages

In this lab we will work with the `tidyverse` and `mosaicData` packages.

```{r}
library(tidyverse) 
library(mosaicData) 
```

Note that these packages are also loaded in your R Markdown document.

# The data

Today's data comes from a study of conducted in Whickham, England. In this study, the researchers recorded each participant's age, smoking status at the start of the study, and their health outcome 20 years later. 

The data is in the `mosaicData` package. You can load it with

```{r load-data}
data(Whickham)
```

Take a peek at the codebook with

```{r eval=FALSE}
?Whickham
```


# Exercises

**Write all R code according to the style guidelines discussed in class.** Be
especially careful about staying within the 80 character limit, as demonstrated 
by your lab TAs! Don't forget to name your code chunks, and commit often!

1. How many observations are in this dataset? What does each observation 
   represent?

1. How many variables are in this dataset? What type of variable is each? 
   Display each variable using an appropriate visualization.
   
1. What would you expect the relationship between smoking status and health outcome to be?

1. Create a visualization depicting the relationship between smoking status and health outcome.

1. Calculate the conditional probabilities of death for each smoking status, only reporting probabilities for the `outcome` of Dead. Briefly describe the relationship, and evaluate whether or not it is what you expected. Use the visualization from the previous exercise and the conditional probabilities to support your narrative.  

1. Create a new variable called `age_cat` using the following scheme:

    - `age <= 44 ~ "18-44"`
    - `age > 44 & age <= 64 ~ "45-64"`
    - `age > 64 ~ "65+"`

1. Re-create the visualization from Exercise 4, this time faceting by `age_cat`.


1. Extend the table from Exercise 5 by breaking it down by age category. 

1. Compare the visualization from Exercise 7 and the table from Exercise 8 to what you previously observed in Exercises 4 and 5. What changed, and what might explain the change? Use the table you calculated in Exercise 8 to support your narrative. 

# Submission

Knit to PDF to create a PDF document. Knit and commit all remaining changes, 
and push your work to GitHub. Make sure all files are updated on your GitHub 
repo.

Please only upload your PDF document to Gradescope. Associate the "Overall"
graded section with the first page of your PDF, and mark where each answer is 
to the exercises. If any answer spans multiple pages, then mark all pages.

<br> 

<hr> 


*This lab was adapted from [Data Science in a Box](https://datasciencebox.org).*
