---
title: "HW 02: Regression"
date: "Due:  11:59pm"
author: "Total: ? points"
output: html_document
editor_options: 
  chunk_output_type: console
---

```{r setup, include=FALSE, message = FALSE, warning = FALSE, echo = FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(caret)
```



## Introduction

In class, we learned about the coefficient of determination $R^2$. In linear regression models, $R^2$ quantifies that proportion of variation in the response that is explained by the predictors. 


## Exercise 1
What is "good" R2? Depends on context.

## Exercise 2
Correlation: what even is it

## Exercise 3
Relationship to correlation: confirm empirically

```{r}
n <- 200
x <- rnorm(n)
y <- 1 + 0.25*x + rnorm(n, 0, 0.5)
cor(x,y)
lm_fit <- lm(y ~ x)
summary(lm_fit)
sqrt(0.1329)
```

## Exercise 4

Recall from the slides that our ordinary least squares estimate for $\beta_{1}$ is

$$\hat{\beta}_{1} = \frac{\sum_{i=1}^{n}(x_{i} - \bar{x})( y_{i} - \bar{y})}{ \sum_{i=1}^{n}(x_{i} - \bar{x})^2 }$$

While we do not derive it here, the correlation coefficient $r$ is calculated as follows:

$$r = \frac{\sum_{i=1}^{n}(x_{i} - \bar{x})( y_{i} - \bar{y})}{\sqrt{\sum_{i=1}^{n} (x_{i} - \bar{x})^2 (y_{i} - \bar{y})^2 }}$$

a) Re-write $\hat{\beta}_{1}$ as a function of $r$.

b) True or False: $\hat{\beta}_{1}$ has the same sign as $r$. Explain.

c) What is the interpretation of $\beta_{1} = 0$ in terms of the correlation?

## Exercise 5?

Prediction intervals vs confidence intervals



## Submission 

Upload your to Gradescope. Associate the "Overall" graded section with the first page of your PDF, and mark where each answer is  to the exercises. If any answer spans multiple pages, then mark all pages.
