---
title: "Recap: Diamonds"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, echo=FALSE}
knitr::opts_chunk$set(warning=FALSE,
                      message=FALSE)

```

```{r load-packages}
library(tidyverse)
```

- What type of variable is price? 

- Would you expect its distribution to be symmetric, right-skewed, or left-skewed? Why? 

- Make a histogram of the distribution of diamond prices. Does the shape of the distribution match your expectation?

```{r}
#add code here 
```

- How many diamonds of each type of cut are there?

```{r}
#add code here 
```

- Calculate the relative frequency (proportion) of each clarity of diamonds.

```{r}
#add code here 
```

- Calculate the relative frequency of each clarity of diamonds by cut.

```{r}
#add code here 
```

- Plot the relationship between depth and price of only fair cut diamonds.

```{r}
#add code here 
```

- For each type of cut, calculate minimum (`min`), maximum (`max`), mean (`mean`), and median (`median`) price of diamonds of that type.

```{r}
#add code here 
```
