---
title: "Popular TV shows"
subtitle: "Web scraping"
date: "`r Sys.Date()`"
output: html_document
---

Go to RStudio Cloud $\rightarrow$ Web scraping and open the R script `scrape-tvshows.R`. 

```{r warning = FALSE, message = FALSE}
library(rvest)
library(tidyverse)
```

1. Scrape the names, scores, and years of most popular TV shows on IMDB: [http://www.imdb.com/chart/tvmeter](http://www.imdb.com/chart/tvmeter)

```{r}
page <- read_html("http://www.imdb.com/chart/tvmeter")

# a pipeline for each column

titles <- page %>%
  html_nodes(".titleColumn a") %>%
  html_text()

years <- page %>% 
  html_nodes("a+ .secondaryInfo") %>%
  html_text() %>%
  str_replace("\\(", "") %>%
  str_replace("\\)", "") %>%
  as.numeric()

scores <- page %>% 
  html_nodes(".imdbRating") %>%
  html_text() %>%
  str_replace("\n", "") %>% 
  as.numeric()
```

2. Create a data frame called `tvshows` with four variables (rank, name, score, year)

```{r}
tvshows <- tibble(
  name = titles, 
  year = years, 
  score = scores
)

tvshows <- tvshows %>%
  mutate(rank = 1:nrow(tvshows))
```

3. Examine each of the first three TV shows to also obtain

    - Genre
    - Runtime
    - How many episodes so far
    - First five plot keywords
    
Add this information to the `tvshows` data frame you created earlier

```{r}
n <- nrow(tvshows)
tvshows <- tvshows %>%
  mutate(genres = rep(NA, n),
         runtime = rep(NA, n), 
         episodes = rep(NA, n),
         keywords = rep(NA, n)
  ) 
```

The Top 3 TV shows on Tuesday, October 7:

1. American Horror Story
2. Peaky Blinders
3. Unbelievable


```{r}
# 1. American Horror Story
page <- read_html("https://www.imdb.com/title/tt1844624/") 
  
genres <- page %>%
  html_nodes(".txt-block~ .canwrap a") %>%
  html_text() %>%
  str_c(collapse = ", ") %>%
  str_trim()

runtime <- page %>%
  html_nodes("#titleDetails time") %>%
  html_text(trim = TRUE) %>%
  str_replace("min", "") %>%
  as.numeric()

episodes <- page %>%
  html_nodes(".bp_sub_heading") %>%
  html_text() %>%
  str_replace("episodes", "") %>%
  as.numeric()

five_plot_keywords <- page %>%
  html_nodes(".itemprop") %>%
  html_text() %>%
  str_c(collapse = ", ") %>% 
  str_trim()

ahs <- tibble(
  genres = genres, 
  runtime = runtime,
  episodes = episodes, 
  keywords = five_plot_keywords
)

tvshows[1, 5:8] <- ahs
```


```{r}
# 2. Peaky Blinders
page <- read_html("https://www.imdb.com/title/tt2442560/") 
  
genres <- page %>%
  html_nodes(".txt-block~ .canwrap a") %>%
  html_text() %>%
  str_c(collapse = ", ") %>%
  str_trim()

runtime <- page %>%
  html_nodes("#titleDetails time") %>%
  html_text(trim = TRUE) %>%
  str_replace("min", "") %>%
  as.numeric()

episodes <- page %>%
  html_nodes(".bp_sub_heading") %>%
  html_text() %>%
  str_replace("episodes", "") %>%
  as.numeric()

five_plot_keywords <- page %>%
  html_nodes(".itemprop") %>%
  html_text() %>%
  str_c(collapse = ", ") %>% 
  str_trim()

pb <- tibble(
  genres = genres, 
  runtime = runtime,
  episodes = episodes, 
  keywords = five_plot_keywords
)

tvshows[2, 5:8] <- pb
```

```{r}
# Unbelievable
page <- read_html("https://www.imdb.com/title/tt7909970/") 
  
runtime <- page %>%
  html_nodes("#titleDetails time") %>%
  html_text(trim = TRUE) %>%
  str_replace("min", "") %>%
  as.numeric()

episodes <- page %>%
  html_nodes(".bp_sub_heading") %>%
  html_text() %>%
  str_replace("episodes", "") %>%
  as.numeric()

five_plot_keywords <- page %>%
  html_nodes(".itemprop") %>%
  html_text() %>%
  str_c(collapse = ", ") %>% 
  str_trim()

un <- tibble(
  genres = genres, 
  runtime = runtime,
  episodes = episodes, 
  keywords = five_plot_keywords
)
  
tvshows[3, 5:8] <- un
```


```{r}
tvshows %>%
  slice(1:3)
```