---
title: "AE 17: Car prices"
date: "06.28.2021"
output: 
  html_document:
    theme: readable
---
## Announcements 

Exam 2 is moved to Friday, July 9, and there is lab this Friday instead.

**Reminder**: project proposals are due on Friday, July 9 as well. Look to the Project website for more information. A Github repository for your project will soon be made.

## Exercises

```{r load-packages, include = FALSE}
library(tidyverse)
library(broom)
```

The dataset `car_prices.csv` contains attributes of cars offered for sale on 
cars.com in 2017^[The data is from the `Stat2Data` R package.]. 

- `type`: Model (Accord, Maxima, Mazda6)
- `age`: Age of the used car (in years)
- `price`: Price (in thousands of dollars)
- `mileage`: Previous miles driven (in thousands of miles)

```{r load-data, message = FALSE}
car_prices <- read_csv("data/car_prices.csv")
```

## Price and Mileage

Consider a regression model with the response `price` and a single predictor `mileage`. 

### Exercise 1

Create a scatterplot of price and mileage.

### Exercise 2 

Use appropriate functions in `R` to find the fitted model and display the results in tidy format. Write out the equation of the fitted model, then include a visualization of the linear model on the scatterplot from Exercise 1. 

### Exercise 3

Interpret the slope and intercept in the context of the problem.

### Exercise 4

What is the predicted selling price of a car with 50,000 miles?

### Exercise 5

Suppose my friend has a Honda Accord with 225,000 miles. Is it appropriate to use this model to make a prediction for the selling price? Why or why not?

## Price and Type

Consider a regression model with the response `price` and the categorical predictor `type` (Accord, Maxima, Mazda6).


### Exercise 6

Use appropriate functions in `R` to find the fitted model and display the results in tidy format. Write out the equation of the fitted model.

### Exercise 7

Interpret the intercept and slope(s) in the context of the problem.

### Exercise 8

How many terms are in the model for `type`? Is this equal to the number of car types in the data set? If not, briefly explain why the number of terms for `type` in the model differs from the number car types in the data set.

### Exercise 9

How does the average price of Maximas compare to the average price of Accords? 

### Exercise 10

What are possible limitations of your two regression models?