---
title: "AE 13: Bone density"
date: "06.16.2021"
output: 
  html_document:
    theme: readable
---

## Getting started 

Clone the **ae-13** repo on [GitHub](https://github.com/sta199-summer2021) and start a new project in RStudio. Be sure to configure git in the RStudio console, so you can so you can push your results back up to GitHub.

```{r eval=FALSE}
library(usethis)
use_git_config(user.name= "github username", user.email="your email")
```

```{r, warning = FALSE, message  = FALSE}
library(tidyverse)
library(infer)
```

## CLT

Suppose the mean bone density for a 65-year old woman is normally 
distributed with mean 809 mg/cm3 and standard deviation of 140 mg/cm3 (these are actually pretty accurate numbers).

### Exercise 1

What bone densities correspond to the 5th, 10th, 50th, and 99th percentiles of this distribution?

### Exercise 2

The densities of three types of wood are below:

- Plywood: 540 mg/cubic centimeter
- Pine: 600 mg/cubic centimeter
- Mahogany: 710 mg/cubic centimeter

What is the probability that a random 65-year old woman has bones less
dense than each of these three woods?

### Exercise 3

What is the probability that, for a *group* of 10 randomly sampled 65-year old women, the mean bone density of the *group* is less dense than each of these three woods?

### Exercise 4

Explain the differences between your answers in Exercise 2 and 3.
