---
title: "AE 15: Parkinson's disease"
date: "06.21.2021"
output: 
  html_document:
    theme: readable
---

## Announcements 

- No class this Wednesday and Thursday, but still lab on Friday!


### Clone a repo + start a new project

- Clone the **ae-15** repo and start a new project in RStudio. 

- Configure git by running the code below in the console. Be sure to fill in your GitHub username and email associated with your Github account. 

```{r eval=FALSE}
library(usethis)
use_git_config(user.name= "your github username", user.email="your email")
```

## Load data and packages 

```{r load-packages-data, eval = FALSE}
library(tidyverse)
library(infer)
parkinsons <- read_csv("data/parkinsons.csv")
```

### Exercise 1

Construct and interpret a 95% confidence interval for the difference in mean shimmer (in dB) between the PD and non-PD individuals using a simulation-based approach.


### Exercise 2

Repeat Exercise 1 using a CLT-based approach.

### Exercise 3

Is there evidence that there is a difference in shimmer between these two groups? Perform a formal statistical hypothesis test using a simulation-based method to answer this question. Include the hypotheses, null distribution, p-value and conclusion stated in the context of the research problem. 


### Exercise 4

Repeat Exercise 3 using the Central Limit Theorem.