---
title: "AE 19: Number of trail users"
subtitle: "MLR: Inference + conditions"
date: "07.01.2021"
output: 
  html_document:
    theme: readable
---
## Announcements 

No class next Monday, July 5. Enjoy your long weekend! 

Monday office hours are cancelled, and rescheduled to Tuesday, July 6 from 9:00-10:30am.

### Project proposal due in a week from tomorrow (July 9)

- Write the proposal in the `proposal.Rmd` file in the project repo, and upload your data into the `data` folder. See the [project description](https://sta199-summer2021.github.io/website/project/project.html) for more information.

- Proposal should include:
   - a description of the data
   - defintion of the variables you intend to explore
   - EDA
   - at least two questions that you are interested in answering using the data


## Exercises 

```{r load-packages, include = FALSE}
library(tidyverse)
library(broom)
```

The Pioneer Valley Planning Commission collected data in Florence, MA for 90 days from April 5 to November 15, 2005 using a laser sensor, with breaks in the laser beam recording when a rail-trail user passed the data collection station. The dataset is called `RailTrail`, and we will be examining the following variables:

- `hightemp`: daily high temperature (in degrees Fahrenheit)
- `volume`: estimated number of trail users that day (number of breaks recorded)
- `fall`: 1: season is Fall, 0: season is not Fall

```{r warning = F}
library(mosaicData)
data(RailTrail)
```

### Exercise 1

Fit a linear model using the daily high temperature to predict the estimated number of trail users. Calculate the $R^2$ and $Adj. R^2$ for the model. 

### Exercise 2

Fit a linear model using the daily high temperature and whether it is the fall season to predict the estimated number of trail users. Calculate the $R^2$ and $Adj. R^2$ for the model. 

### Exercise 3

Which model has the higher $R^2$? Which model has the higher $Adj. R^2$? Which model is a better fit for the data? 


### Exercise 4

Let's use the model from Exercise 1. Display the output for the model from Exercise 1 with the 90% confidence interval for the model coefficients.

Interpret the coefficient of `hightemp` in the context of the data. 

### Exercise 5

Interpret the 90% confidence interval for `hightemp` in the context of the data. 

### Exercise 6

Does the intercept have a reasonable interpretation? If so, interpret the intercept in the context of the data. Otherwise, explain why not. 

### Exercise 7

Conduct a hypothesis test to assess if there is a linear relationship the high temperature and number of trail users. State the null and alternative hypotheses, the test statistic, p-value, and conclusion in the context of the data.

### Exercise 8

Is the confidence interval consistent with the conclusion from the hypothesis test in the previous exercise? Briefly explain why or why not. 

### Exercise 9

Let's check the model conditions. Create the diagnostic plots use them to determine if the four model conditions are met.