---
title: "AE 14: Pokemon"
date: "06.17.2021"
output: 
  html_document:
    theme: readable
---

## Announcements 

- There will be no class next Wednesday and Thursday, June 23-24. However, there
will still be lab on Friday, June 25. 

- Becky's Wednesday office hours will be cancelled due to travel, but are re-scheduled to Thursday, June 24 at 3:00-4:30pm (and by appointment, as always).

## Comparing CLT- & simulation- based inference

What is similar, and what is different, between CLT based test of means vs. simulation based test?

## Getting started 

Clone the **ae-14** repo on [GitHub](https://github.com/sta199-summer2021) and start a new project in RStudio. Be sure to configure git in the RStudio console, so you can so you can push your results back up to GitHub.

```{r eval=FALSE}
library(usethis)
use_git_config(user.name= "github username", user.email="your email")
```

## CLT-based inference with Pokemon

```{r, warning = FALSE, message  = FALSE}
library(tidyverse)
library(infer)
```

We will be using the `pokemon` dataset, which contains information about 45 randomly selected Pokemon (from all generations). You may load in the dataset with the following code: 

```{r load-data, warning = FALSE, message = FALSE, eval = F}
pokemon <- read_csv("data/pokemon.csv")
```

**Use the infer package to answer the following exercises.**

### Exercise 1

Construct and interpret a 95% confidence interval for the mean height in meters (`height_m`) of Pokemon species by using the Central Limit Theorem. 

### Exercise 2

The average height of a human is 1.65 meters. Evaluate whether a randomly selected Pokemon species has a *different* mean height by using the Central Limit Theorem. 

In doing so, state your null and alternative hypotheses, the
distribution of your test statistic under the null hypothesis, your p-value, decision, and conclusion in context of the research problem.

### Exercise 3

Now evaluate whether a randomly-selected Pokemon species has a *lower* mean height by using the Central Limit Theorem. In doing so, state your null and alternative hypotheses, the distribution of your test statistic under the null hypothesis, your p-value, decision, and conclusion in context of the research problem.

### Exercise 4

Suppose the true mean height among Pokemon species is 1.4 meters. In your conclusions from Exercises 2 and 3, did you make the correct decision, a Type 1 error, or a Type 2 error? Explain.
