---
title: "AE 15: Parkinson's disease"
date: "06.21.2021"
output: 
  html_document:
    theme: readable
---

## Announcements 

- No class this Wednesday and Thursday, but still lab on Friday!
- HW 02 will be released this Wednesday, and is due on Wednesday, June 30

### Exam 2

Proposed new schedule: originally, Exam 02 would have been assigned next Friday, July 2. However,
in order to give you a bit more of a break in between exams, allow you to enjoy your holiday,
and get feedback in time on HW 02, I am considering moving Exam 02 the following weekend, Friday July 9.

Because the syllabus originally stated the date of Exam 02 as July 2, if this new date 
does not work for **at least one person**, the exam date will go back to the original.
**Please email Becky by Wednesday, June 23 if the proposed new date does not work**.



### Clone a repo + start a new project

- Clone the **ae-15-real** repo and start a new project in RStudio. 

- Configure git by running the code below in the console. Be sure to fill in your GitHub username and email associated with your Github account. 

```{r eval=FALSE}
library(usethis)
use_git_config(user.name= "your github username", user.email="your email")
```

## Load data and packages 

```{r load-packages-data, eval = FALSE}
library(tidyverse)
library(infer)
parkinsons <- read_csv("data/parkinsons.csv")
```

### Exercise 1

Construct and interpret a 95% confidence interval for the difference in mean shimmer (in dB) between the PD and non-PD individuals using a simulation-based approach.


### Exercise 2

Repeat Exercise 1 using a CLT-based approach.

### Exercise 3

Is there evidence that there is a difference in shimmer between these two groups? Perform a formal statistical hypothesis test using a simulation-based method to answer this question. Include the hypotheses, null distribution, p-value and conclusion stated in the context of the research problem. 


### Exercise 4

Repeat Exercise 3 using the Central Limit Theorem.