---
title: "AE 11: Hypothesis Testing"
subtitle: "Part 1"
date: "06.09.2021"
output: 
  html_document:
    theme: readable
---

```{r echo = F, warning = FALSE, message = FALSE}
library(tidyverse)
library(infer)
```



## Hypothesis testing review

$$H_0: p = 0.1 \text{ vs }H_a: p < 0.1$$

```{r, echo = FALSE}
organ_donor <- tibble(
  outcome = c(rep("complication", 3), rep("no complication", 59))
)
```

**Generate the null distribution**

```{r}
null_dist <- organ_donor %>%
  specify(response = outcome, success = "complication") %>% #<<
  hypothesize(null = "point", 
              p = c("complication" = 0.10, "no complication" = 0.90) 
              ) %>% 
  generate(reps = 100, type = "simulate") %>% 
  calculate(stat = "prop")
```

**Calculate p-value**

```{r}
null_dist %>%
  filter(stat <= (3/62)) %>%
  summarise(p_value = n()/nrow(null_dist))
```


## Clone a repo + start a new project

Clone the **ae-11** repo on [GitHub](https://github.com/sta199-summer2021) and start a new project in RStudio. Be sure to configure git in the RStudio console, so you can so you can push your results back up to GitHub.

```{r eval=FALSE}
library(usethis)
use_git_config(user.name= "github username", user.email="your email")
```

We will be using the `asheville` dataset. You may load in the dataset with the
following code (be sure to set `eval` to be `TRUE` in the following R chunk!):

```{r load-packages, warning = FALSE, message = FALSE}
library(tidyverse)
library(infer)
```


```{r load-data, message = FALSE, echo = FALSE}
asheville <- read_csv("data/asheville.csv")
```

### Exercise 1 (Zoom poll)

Suppose you are interested in whether the mean price per guest per night is actually less than $80. Choose the correct null and alternative hypotheses.

a. $H_0: \mu = 80 \text{ vs. }H_a: \mu \neq 80$
b. $H_0: \mu = 80 \text{ vs. }H_a: \mu < 80$
c. $H_0: \mu = 80 \text{ vs. }H_a: \mu > 80$
d. $H_0: \bar{x} = 80 \text{ vs. }H_a: \bar{x}  \neq 80$
e. $H_0: \bar{x} = 80 \text{ vs. }H_a: \bar{x}  < 80$
f. $H_0: \bar{x} = 80 \text{ vs. }H_a: \bar{x}  > 80$


### Exercise 2

Let's use simulation-based methods to conduct the hypothesis test specified in Exercise 1. We'll start by generating the null distribution. 

Fill in the code and uncomment the lines below to generate the null distribution. 

```{r set-seed}
set.seed(092120)
```


```{r ex-2, eval = F}
#null_dist <- asheville %>%
  #specify(response = ______) %>%
  #hypothesize(null = ______, mu = ______) %>%
  #generate(reps = 100, type = "bootstrap") %>%
  #calculate(stat = _____)
```

### Exercise 3

Fill in the code and uncomment the lines below to calculate the p-value using the null distribution from Exercise 2.

```{r calc-mean}
mean_ppg <- asheville %>% 
  summarise(mean_ppg = mean(ppg)) %>%
  pull()
```

```{r eval = F}
#null_dist %>%
  #filter(______) %>%
  #summarise(p_value = ______)
```

## Exercise 4 

Use the p-value to make your conclusion using a significance level of 0.05. Remember, the conclusion has 3 components 

- How the p-value compares to the significance level 
- The decision you make with respect to the hypotheses (reject $H_0$ /fail to reject $H_0$)
- The conclusion in the context of the research problem 

### Exercise 5

Suppose you are interested in whether the **median** price per guest per night is
equal to or less than $80. Carry out a similar analysis to that
undertaken in Exercises 1 - 4 to test these hypotheses. 
