---
title: "AE 16: Two-sample Inference"
date: "06.21.2021"
output: 
  html_document:
    theme: readable
    toc: true
    toc_depth: 2
---

### Announcements 

Happy STA 199 break! 

- Tomorrow's lab: Mid-semester feedback + Project proposal work day
- No class Wednesday or Thursday
- Becky's Wednesday office hours are cancelled, and rescheduled to Thursday 3:00-4:30


Let's look at the diamonds dataset again, this time using a subset of 10000 observations.


```{r message = F, warning = F, echo = F}
library(tidyverse)
library(infer)
data(diamonds)
set.seed(100)
diamonds_sub <- diamonds %>%
  sample_n(10000) %>%
  mutate(cut = as.character(cut),
         color = as.character(color),
         clarity = as.character(clarity))
```

### Exercise 1

Is there evidence that
the average price of Good diamonds is different from the average price of Ideal diamonds?

Evaluate this question using a simulation based approach. State the null and alternative hypotheses, the p-value, and conclusion in the context of the research problem.  

```{r}

```

### Exercise 2

Answer the same question above using a 95% confidence interval 
obtained using a CLT-based approach. Don't forget to check your assumptions. How
does your answer here relate to your conclusion for Ex. 1?

```{r}

  
```

### Exercise 3

Assess whether there is a relationship between the color of the diamond and its 
clarity using a hypothesis test at the $\alpha = 0.05$ level. State your hypothesis, 
provide the test-statistic and its distribution, and evaluate your p-value and 
decision in the context of the problem.

