---
title: "AE 04: Data Wrangling"
author: "Your Name"
date: "05.20.2021"
output: 
  html_document:
    theme: readable
---


## Clone a repo + start a new project

Go to the `ae-04-[GITHUB USERNAME]` repo, clone it, and start a new project in RStudio. 

## Configure git

Run the following code to configure Git. Fill in your GitHub username and the email address associated with your GitHub account. 

```{r eval=FALSE}
library(usethis)
use_git_config(user.name= "your github username", user.email="your email")
```

## Practice with data wrangling

```{r load-packages, warning = FALSE, message = FALSE}
library(tidyverse)
```

The data set contains all North Carolina bike crash data from 2007-2014. The data was originally downloaded on September 6, 2018 and is available in the dsbox R package. 

Use the code below to load the data from the .csv file

```{r load-data, warning = FALSE, message = FALSE}
ncbikecrash <- read_csv("data/ncbikecrash.csv")
```


### Part 1

**Display the first 10 rows of the ncbikecrash data set.**

```{r}
```


### Part 2 

**Filter for bike crashes in Durham County where the driver age group is 0 to 19.**

```{r}
```

### Part 3

**On which day of the week did the most bike crashes occur? Use the `count` function to get the number of bike crashes for each day (`crash_day`). Display the results in descending order.**

```{r}
```

### Part 4

**Calculate the average estimated crash hour, `crash_hour` (<u>not</u> driver_est_speed) based on whether the crash was a hit and run (`hit_run`).**

```{r}
```