---
title: "AE 07: Alinity COVID-19 Test"
name: "Your name"
date: "05.27.2021"
output: 
  html_document:
    theme: readable
---

## Clone a repo + start a new project

Go to the `ae-07-[GITHUB USERNAME]` repo, clone it, and start a new project in RStudio. 

## Configure git

Run the following code to configure Git. Fill in your GitHub username and the email address associated with your GitHub account. 

```{r eval=FALSE}
library(usethis)
use_git_config(user.name = "your github username", user.email ="your email")
```

```{r load-packages, warning = FALSE, message = FALSE}
library(tidyverse)
```

**Use R as a calculator for the exercises in this AE.** 

## Warm-up

Suppose in a class that 30% of students were Pratt students and 70% were 
Trinity. 10% of Pratt students are vegetarian and 5% of Trinity students are 
vegetarian. 

### Exercise 1

What is the probability that a randomly selected student in this class is 
vegetarian?

```{r ex-1}
```

### Exercise 2

Given that a student is vegetarian, what is the probability that they are in 
Pratt?

```{r ex-2}
```

## The Abbott Alinity Test

The global coronavirus pandemic illustrates the need for accurate testing of 
COVID-19, as its extreme infectivity poses a significant public health threat. 
Due to the time-sensitive nature of the situation, the FDA has enacted emergency 
authorization of a number of serological tests for COVID-19. Full details of 
these tests may be found on its website 
[here](https://www.fda.gov/medical-devices/coronavirus-disease-2019-covid-19-emergency-use-authorizations-medical-devices/eua-authorized-serology-test-performance).

The Abbott Alinity test has an estimated sensitivity of 100% and specificity of 
99%. Recall from lecture: $D$ is the event that a patient has the disease, and let $T$ is the event that the test is positive for that disease.

- Prevalence: $P(D)$
- Sensitivity: $P(T | D)$
- Specificity: $P(T^c | D^c)$
- Positive predictive value: $P(D | T)$
- Negative predictive value: $P(D^c | T^c)$

### Exercise 3

Suppose the prevalence of COVID-19 in the general population is 2%. What are 
the positive and negative predictive values of the Abbott Alinity test?

```{r ex-3}
```

### Exercise 4

Now suppose the prevalence of COVID-19 in the general population is 8%. Repeat
the same calculations as in Exercise 3. Can you reuse code from Exercise 3
somehow?

```{r ex-4}
```

### Exercise 5

In the context of COVID-19, would you prefer to have a test with perfect 
sensitivity and good specificity or would you rather have a test with perfect 
specificity and good sensitivity? Explain your answer.
