---
title: "AE 22: ANOVA"
date: "07.12.2021"
output: 
  html_document:
    theme: readable
---

```{r echo = F}
knitr::opts_chunk$set(warning = F, message = F)
```

```{r echo  = F}
library(tidyverse)
library(broom)
library(infer)
```

## Announcements


## Exercises 

Let's look at data relating student test scores in various subjects to factors such as
parent background and test preparation. The data are from [Kaggle](https://www.kaggle.com/spscientist/students-performance-in-exams). We'll use the following variables in the analysis: 

- `math` - score on math test (between 0 and 100)
- `parent_educ` - highest level of parental education: associate's degree, bachelor's degree, high school,
master's degree, some college, some high school


```{r}
scores <- read.csv("data/student_performance.csv")
```

### Exercise 1
Visualize the relationship between math score and parent education. Describe any patterns or relationships you see.


### Exercise 2
Write the ANOVA hypotheses to test the relationship between parent education and math test score.


### Exercise 3
Check the assumptions of ANOVA by (1) normality and (2) homoscedastic variance. Assuming the observations
are independent, should we move forward with the ANOVA testing?


### Exercise 4
Regardless of what you concluded in Exercise 3, perform the hypothesis test using the ANOVA model and a significance level of $\alpha = 0.05$. State the value of the test statistic, as well the distribution it follows, assuming $H_{0}$ true. 


### Exercise 5
Should you proceed with any more testing? Explain your decision.


### Exercise 6
If we were to proceed with testing and conducted all pairwise tests, we would conduct 15 t-tests. 
What would be the Bonferroni-correct significance level we would use for each test?


## Exercise 7
Pick one pairwise hypothesis and perform a t-test using the significance level from Exercise 6. **Hint**: use the `t_test()` function, filtering for the two levels of parent education that you are testing between.

