---
title: "AE 18: Paris paintings"
date: "06.30.2021"
output: 
  html_document:
    theme: readable
---

## Announcements

Your project Github repositories have been created. Reminder that your project proposal is due next Friday, July 9 at 11:59pm. Use the file `project-proposal.Rmd` to write your proposal.


## Exercises 

```{r load-packages, include = FALSE}
library(tidyverse)
library(broom)
```

```{r load-data, message = FALSE}
paintings <- read_csv("data/paris_paintings.csv", 
                            na = c("n/a", "", "NA"))
```

We will be looking at the Paris Paintings data set in today's application exercise. We'll primarily focus on the variables: 

- `Height_in`: Height (in inches)
- `Width_in`: Width (in inches)
- `landsAll`: If any type of landscape is mentioned (either lands_sc, lands_figs, or lands_ment)



### Exercise 1

Create a scatterplot to visualize the relationship between Height and Width. Color the points based on the whether the painting has any landscape elements. *Note: Be sure to make `landsAll` a factor, so R treats it as categorical variable.*

### Exercise 2

Based on your scatterplot, does the relationship between Height and Width differ between paintings with landscape elements and those without? Briefly explain.  

### Exercise 3

Fit a main effects model using width and whether the painting has landscape elements to predict the height. 

### Exercise 4

Using your model from the previous exercise, 

- Interpret the intercept. 
- Interpret the coefficient (slope) of `Width_in`. 
- Interpret the coefficient (slope) of `landsAll`.

### Exercise 5

Write the equation of the model for 

- paintings with no landscape elements. 
- paintings with landscape elements. 

How do the slopes compare between the two models? How do the intercepts compare? 

### Exericse 6

Now let's consider an interaction term. Fit a linear model using width, whether the painting has landscape elements, and the interaction between the two variables to predict the height. 

### Exercise 7

Write the equation of the model for 

- paintings with no landscape elements. 
- paintings with landscape elements. 

How do the slopes compare between the two models? How do the intercepts compare? 

### Exercise 8

Now let's see how well each model fits the data. Use the `glance` function to calculate $R^2$ and Adj. $R^2$ for the model fit in Ex. 3 and the one fit in Ex. 6. 

### Exercise 9

Based on your answer to the previous exercise, which model is a better fit for the data? Briefly explain.

### Exercise 10

Interpret the $R^2$ for the model chosen in the previous exercise.