---
title: "AE 13: Bone density"
date: "06.14.2021"
author: "Your name"
output: 
  html_document:
    theme: readable
---

## Announcements 

- There will be no class next Wednesday and Thursday, June 23-24. However, there
will still be lab on Friday, June 25. Use this time to work on HW 02. 

- Becky's Wednesday office hours will be cancelled due to travel, but are re-scheduled
to Thursday, June 24 at 3:00-4:30pm (and by appointment, as always).

### Exam

## Comparing CLT- & simulation- based inference

What is similar, and what is different, between CLT based test of means vs. simulation based test?

## Getting started 

Clone the **ae-13** repo on [GitHub](https://github.com/sta199-summer2021) and start a new project in RStudio. Be sure to configure git in the RStudio console, so you can so you can push your results back up to GitHub.

```{r eval=FALSE}
library(usethis)
use_git_config(user.name= "github username", user.email="your email")
```

```{r, warning = FALSE, message  = FALSE}
library(tidyverse)
library(infer)
```

## CLT

Suppose the mean bone density for a 65-year old woman is normally 
distributed with mean 809 mg/cm3 and standard deviation of 140 mg/cm3 (these are actually pretty accurate numbers).

### Exercise 1

What bone densities correspond to the 5th, 10th, 50th, and 99th percentiles of this distribution?

### Exercise 2

The densities of three woods are below:

- Plywood: 540 mg/cubic centimeter
- Pine: 600 mg/cubic centimeter
- Mahogany: 710 mg/cubic centimeter

What is the probability that a random 65-year old woman has bones less
dense than each of these three woods?

### Exercise 3

What is the probability that, for a *group* of 10 randomly sampled 65-year old women, the mean bone density of the *group* is less dense than these three woods?

### Exercise 4

Explain the differences between your answers in Exercise 2 and 3.


## CLT-based inference with Pokemon

We will be using the `pokemon` dataset, which contains information about 45 randomly selected Pokemon (from all generations). You may load in the dataset with the following code: 

```{r load-data, warning = FALSE, message = FALSE, eval = T}
pokemon <- read_csv("data/pokemon.csv")
```

**Use the infer package to answer the following exercises.**

### Exercise 5

Construct and interpret a 95% confidence interval for the mean height in meters (`height_m`) of Pokemon species by using the Central Limit Theorem. 

### Exercise 6

The average height of a human is 1.65 meters. Evaluate whether a randomly selected Pokemon species has a *different* mean height by using the Central Limit Theorem. 

In doing so, state your null and alternative hypotheses, the
distribution of your test statistic under the null hypothesis, your p-value, decision, and conclusion in context of the research problem.

### Exercise 7

Now evaluate whether a randomly-selected Pokemon species has a *lower* mean height by using the Central Limit Theorem. In doing so, state your null and alternative hypotheses, the distribution of your test statistic under the null hypothesis, your p-value, decision, and conclusion in context of the research problem.

### Exercise 8

Suppose the true mean height among Pokemon species is 1.4 meters. In your conclusions from Exercises 6 and 7, did you make the correct decision, a Type 1 error, or a Type 2 error? Explain.
