---
title: "Teamwork & Project"
author: ""
date: "10.11.19"
output:
  xaringan::moon_reader:
    mathjax: "https://cdn.bootcss.com/mathjax/2.7.1/MathJax.js?config=TeX-MML-AM_HTMLorMML"
    css: "sta199-slides.css"
    logo: img/sta199-sticker-icon.png
    lib_dir: libs
    nature:
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r setup, include=FALSE}
# R options
options(
  htmltools.dir.version = FALSE, # for blogdown
  show.signif.stars = FALSE,     # for regression output
  warm = 1
  )
# Set dpi and height for images
library(knitr)
opts_chunk$set(fig.height = 2.35, dpi = 300) 
# ggplot2 color palette with gray
color_palette <- list(gray = "#999999", 
                      salmon = "#E69F00", 
                      lightblue = "#56B4E9", 
                      green = "#009E73", 
                      yellow = "#F0E442", 
                      darkblue = "#0072B2", 
                      red = "#D55E00", 
                      purple = "#CC79A7")
# For nonsese...
library(emo)
htmltools::tagList(rmarkdown::html_dependency_font_awesome())
```

```{r echo=FALSE, message=FALSE, warning=FALSE}
library(tidyverse)
library(rvest)
```

## Agenda

1. Teamwork Activity

2. Introducing the Project

---

class: middle, center

## Working on Teams

---

### Why do we work on teams? 

- To learn from one another!

--

- Teamwork is a highly valued skill. 
    - In a [survey by the Association of American Colleges & Universities](https://www.aacu.org/leap/public-opinion-research/2015-survey-falling-short), 83% of the employers rated "The ability to work effectively with others in teams" as very important. 


---

### Teamwork Reflection 

Take out a piece of paper. 

--

1. Reflect on your role on the team thus far. Write 1 - 3 things you can do to be an even better team member. 

--

2. Each group member...
    - Share what you wrote with your team. 
    - **All other team members**: Say one way you will help your teammate do the things they wish to improve. 

- Next team feedback you will be asked how much each team member has improved on their goals. This will be shared feedback. 


---

### Project

- **Instructions**: https://www2.stat.duke.edu/courses/Fall19/sta199.001/project/project.html

--

- **Today's Assignment: due Wed, Oct 15**
    - Identify 3 - 4 topics of interest
    - Identify datasets related to at least two of the topics
    - Write your responses in the `topic-ideas` file in the `project` repo on GitHub
    

    




